---
title: "Meta analysis"
date: today
format: html
code-tools: true
code-fold: true
embed-resources: false
editor_options: 
  chunk_output_type: console
---

```{r}
#| code-summary: Load packages and setup
#| message: false
#| warning: false

suppressPackageStartupMessages({
  suppressWarnings({
    library(tidyverse)
    library(lme4)
    library(metafor)
  })
})
```

```{r}
#| code-summary: Load dataset
match_injury <- readRDS(here::here("output/match-played-seleced-team.rds")) |> 
  summarize(
    across(c(n_match, n_injury), \(x) sum(x, na.rm = TRUE)),
    .by = c(league, league_tag, team, team_tag, season)
  ) |> 
  mutate(
    match_exposure = n_match * 11 * 1.5,
    match_injury_rate = (n_injury / match_exposure) * 1000
  ) |> 
  mutate(season = str_extract(season, "\\d+") |> as.numeric()) |> 
  separate_wider_regex(league, c(country = ".*?", ". ", league = ".*"))
  
match_injury |> glimpse()
```

## Injury Rate (2019–2024)

::: panel-tabset

### By league and season

```{r}
#| code-summary: Plot match injury rate by league and season
#| fig-width: 8
#| out-width: 100%
#| fig-asp: 0.9

match_injury |> 
  filter(between(season, 2021, 2024)) |> 
  ggplot(aes(n_match, match_injury_rate)) +
  geom_point(shape = 21, fill = "whitesmoke", stroke = 1) +
  geom_smooth(method = "lm", formula = "y ~ x", fullrange = TRUE) +
  facet_grid(cols = vars(season), rows = vars(league)) +
  scale_color_brewer(palette = "Set1") +
  coord_cartesian(xlim = c(20, 75)) +
  theme(
    legend.position = "bottom",
    legend.justification = "left"
  ) +
  labs(
    x = "Number of match",
    y = "Injury Rate (#injury/#match exposure) * 1000"
  )
```

### By season

```{r}
#| code-summary: Plot match injury rate by season
#| fig-width: 7
#| out-width: 100%
#| fig-asp: 0.8

match_injury |> 
  filter(between(season, 2019, 2024)) |>
  ggplot(aes(n_match, match_injury_rate)) +
  geom_point(shape = 21, fill = "whitesmoke", stroke = 1) +
  geom_smooth(method = "lm", formula = "y ~ x", fullrange = TRUE) +
  facet_wrap(facet = vars(season)) +
  scale_color_brewer(palette = "Set1") +
  coord_cartesian(xlim = c(20, 75)) +
  theme(
    legend.position = "bottom",
    legend.justification = "left"
  ) +
  labs(
    x = "Number of match",
    y = "Injury Rate (#injury/#match exposure) * 1000"
  )
```

### Overall

```{r}
#| code-summary: Plot match injury rate by season
#| fig-width: 7
#| out-width: 100%
#| fig-asp: 0.8

match_injury |> 
  filter(between(season, 2019, 2024)) |>
  ggplot(aes(n_match, match_injury_rate)) +
  geom_point(
    shape = 21, fill = "whitesmoke", stroke = 1, 
    position = position_jitter(width = 2, height = 2)
  ) +
  geom_smooth(method = "lm", formula = "y ~ x", fullrange = TRUE) +
  scale_color_brewer(palette = "Set1") +
  coord_cartesian(xlim = c(20, 75)) +
  theme(
    legend.position = "bottom",
    legend.justification = "left"
  ) +
  labs(
    x = "Number of match",
    y = "Injury Rate (#injury/#match exposure) * 1000"
  )
```

:::

## Modelling

### Joint Analysis

```{r}
#| code-summary: Generalized mix-effect model - league:team as random factor

match_injury_joint_model <- match_injury |> 
  mutate(
    season_z = (season - mean(season)) / sd(season),
    match_played_z = (n_match - mean(n_match)) / sd(n_match)
  ) |> 
  lme4::glmer(
    n_injury ~ match_played_z + season_z + league + (1 | league:team),
    data = _,
    family = poisson(link = "log"),
    offset = log(match_exposure)
  )
summary(match_injury_joint_model)
```

### Separate model for each league

```{r}
#| code-summary: Generalized mix-effect model by league - team as random factor

match_injury_league_models <- match_injury |> 
  mutate(
    season_z = (season - mean(season)) / sd(season),
    match_played_z = (n_match - mean(n_match)) / sd(n_match)
  ) |> 
  group_by(league) |>
  group_map(~lme4::glmer(
    n_injury ~ match_played_z + season_z + (1 | team),
    data = .x,
    family = poisson(link = "log"),
    offset = log(match_exposure)
  ))
names(match_injury_league_models) <- unique(match_injury$league)
```

::: panel-tabset

```{r}
#| code-summary: Summary of league-specific models
#| results: asis

iwalk(match_injury_league_models, function(.mdl, .name) {
  cat(glue::glue("#### {.name}"), "\n\n")
  cat("```txt\n")
  cat(paste(capture.output(summary(.mdl)), collapse = "\n"))
  cat("\n```\n\n")
  cat("\n\n")
})
```

:::


## Meta Analysis

```{r}
#| code-summary: Gather coefficients from each league specific model

league_models_coef <- match_injury_league_models |> 
  map_dfr(broom.mixed::tidy, .id = "league")
```

### Fixed effect model

::: panel-tabset

#### Summary

```{r}
#| code-summary: Meta analysis using fixed-effect method
#| results: asis

meta_analysis_fe <- league_models_coef |> 
  filter(is.na(group)) |> 
  group_by(term) |> 
  group_map(function(.term, .name) {
    metafor::rma.uni(
      yi = estimate,
      sei = std.error,
      data = .term,
      method = "FE",
      slab = league
    )
  }) |> 
  setNames(head(gsub("[\\(\\)]", "", unique(league_models_coef$term)), -1))

cat("::: panel-tabset\n\n")
iwalk(meta_analysis_fe, function(.mdl, .name) {
  cat(glue::glue("#### {.name}"), "\n\n")
  
  cat("```txt\n")
  cat(paste(capture.output(summary(.mdl)), collapse = "\n"))
  cat("\n```\n\n")
  
  cat("\n\n")
})
cat(":::\n\n")
```

#### Plot

```{r}
#| code-summary: Estimates from meta analysis using random-effect
#| results: asis
#| fig-width: 9
#| out-width: 100%
#| fig-asp: 0.7

cat("::: panel-tabset\n\n")
iwalk(meta_analysis_fe, function(.mdl, .name) {
  cat(glue::glue("#### {.name}"), "\n\n")
  metafor::forest.rma(x = .mdl, showweights = TRUE)
  cat("\n\n")
})
cat(":::\n\n")
```

:::

### Random effect model

::: panel-tabset

#### Summary

```{r}
#| code-summary: Meta analysis using random-effect method
#| results: asis

meta_analysis_re <- league_models_coef |> 
  filter(is.na(group)) |> 
  group_by(term) |> 
  group_map(function(.term, .name) {
    metafor::rma.uni(
      yi = estimate,
      sei = std.error,
      data = .term,
      method = "REML",
      slab = league
    )
  }) |> 
  setNames(head(gsub("[\\(\\)]", "", unique(league_models_coef$term)), -1))

cat("::: panel-tabset\n\n")
iwalk(meta_analysis_re, function(.mdl, .name) {
  cat(glue::glue("#### {.name}"), "\n\n")
  
  cat("```txt\n")
  cat(paste(capture.output(summary(.mdl)), collapse = "\n"))
  cat("\n```\n\n")
  
  cat("\n\n")
})
cat(":::\n\n")
```

#### Plot

```{r}
#| code-summary: Estimates from meta analysis using random-effect
#| results: asis
#| fig-width: 9
#| out-width: 100%
#| fig-asp: 0.7

cat("::: panel-tabset\n\n")
iwalk(meta_analysis_re, function(.mdl, .name) {
  cat(glue::glue("#### {.name}"), "\n\n")
  
  metafor::forest.rma(
    x = .mdl,
    predstyle = "dist",
    showweights = TRUE
  )
  
  cat("\n\n")
})
cat(":::\n\n")
```

#### BLUPs


```{r}
#| code-summary: Best Linear Unbiased Prediction (BLUP)
#| results: asis
#| fig-width: 9
#| out-width: 100%
#| fig-asp: 0.7

cat("::: panel-tabset\n\n")
iwalk(meta_analysis_re, function(.mdl, .name) {
  cat(glue::glue("#### {.name}"), "\n\n")
  
  cat("**BLUP Estimates**\n")
  
  cat("```txt\n")
  cat(paste(capture.output(blup(.mdl)), collapse = "\n"))
  cat("\n```\n\n")
  
  cat("**Pooled Estimates**\n")
  
  cat("```txt\n")
  cat(paste(capture.output(predict(.mdl)), collapse = "\n"))
  cat("\n```\n\n")
  
  cat("\n\n")
})
cat(":::\n\n")
```


:::

# AI Assisted Report

**Summarizing meta-analysis results**

I need to summarize and analyze the meta-analysis results in a structured way, using both math and plain language. I'll break it down into sections like Introduction, Data & Outcome, Primary Joint Model, League-Specific Models, Meta-analysis of Coefficients, and Practical Interpretation. I'll also touch on limitations and conclusions. I need to explain concepts like the log link, Poisson distribution, and how to compute the Injury Rate Ratio (IRR) using exponentiation, like exp(0.114). I'll keep it concise but complete.
# Injury meta analysis report across the top 5 European leagues

You’ve already done a lot of the heavy lifting with the models—what I’ll do here is turn those results into a coherent story, moving between the math and what it means in football terms.

---

## 1. Data and outcome

The dataset contains:

- **Leagues:** Bundesliga, La Liga, Ligue 1, Premier League, Serie A  
- **Units:** Team–season observations (1,560 rows, 186 teams)  
- **Outcome:**  
  - **`n_injury`** = number of match injuries per team per season  
  - **Exposure:** `match_exposure` (match minutes or match-hours), used as an **offset**  
- **Key predictors:**
  - **`n_match`** (standardized to `match_played_z`): number of matches played  
  - **`season`** (standardized to `season_z`): time trend (2010–2024)  
  - **`league`**: categorical factor  
- **Model family:** Poisson with log link, i.e. modeling **injury counts** with **log(injury rate)** as the linear predictor.

In plain language: you’re modeling **injury rates per unit of match exposure**, adjusting for how many matches a team plays and how late in the time period the season is, with teams nested in leagues.

---

## 2. Joint generalized linear mixed model

### 2.1 Model specification

Joint model:

$$
\log(\lambda_{ij}) = \beta_0 + \beta_1 \text{match\_played\_z}_{ij} + \beta_2 \text{season\_z}_{ij} + \beta_{\text{league}} + u_{\text{league:team}} ,
$$

with

- $\lambda_{ij}$ = expected injury rate for team $j$ in league $i$ per unit exposure  
- Offset: $\log(\text{match\_exposure})$  
- Random intercept: $u_{\text{league:team}} \sim N(0, \sigma^2)$

### 2.2 Key fixed effects (joint model)

From the joint model:

- **Intercept:** $\hat\beta_0 = -4.1875$  
- **Match played (z):** $\hat\beta_1 = 0.1083$  
- **Season (z):** $\hat\beta_2 = 0.5902$  
- **League effects (vs Bundesliga as reference):**  
  - La Liga: $-0.1891$  
  - Ligue 1: $-0.2589$  
  - Premier League: $-0.0288$ (ns)  
  - Serie A: $-0.0511$ (ns)

All p-values for `match_played_z` and `season_z` are \< 0.001.

### 2.3 Interpretation in plain language

- **Baseline injury rate:**  
  The intercept $-4.19$ is the log injury rate at:
  - average number of matches,
  - average season (centered),
  - in the reference league (Bundesliga),
  - for an “average” team.

  Converting to a rate:

  $$
  \text{Rate}_0 = e^{-4.1875} \approx 0.0152
  $$

  So roughly **15 injuries per 1,000 units of exposure** (depending on how exposure is scaled).

- **Effect of number of matches (`match_played_z`):**  
  $$
  \text{IRR}_{\text{match}} = e^{0.1083} \approx 1.11
  $$

  A **1 SD increase in number of matches** is associated with about an **11% higher injury rate**, holding season and league constant. Here as well, possibly affected by better coverage of injuries in more recent years.

- **Effect of season (`season_z`):**  
  $$
  \text{IRR}_{\text{season}} = e^{0.5902} \approx 1.80
  $$

  A **1 SD increase in season (i.e., moving towards more recent years)** is associated with about an **80% higher injury rate**, after adjusting for matches and league. This is a very strong upward time trend.

- **League differences (vs Bundesliga):**
  - La Liga: $e^{-0.1891} \approx 0.83$ → about **17% lower injury rate** than Bundesliga, all else equal.  
  - Ligue 1: $e^{-0.2589} \approx 0.77$ → about **23% lower injury rate**.  
  - Premier League and Serie A: small, non-significant differences vs Bundesliga.

In words: **injury rates are increasing over time**, higher for teams that play more matches, and slightly lower in La Liga and Ligue 1 compared to Bundesliga, with Premier League and Serie A roughly similar to Bundesliga once you adjust for other factors.

---

## 3. League-specific mixed models

You then fit **separate Poisson mixed models by league**, each with:

$$
\log(\lambda_{jt}) = \alpha_0 + \alpha_1 \text{match\_played\_z}_{jt} + \alpha_2 \text{season\_z}_{jt} + u_{\text{team}},
$$

with random intercepts for teams.

### 3.1 Coefficients by league

**Intercepts (log rates):**

- Bundesliga: $-4.18$  
- Ligue 1: $-4.40$  
- Premier League: $-4.46$  
- La Liga: $-4.19$  
- Serie A: $-4.23$

These are all in a fairly narrow band, roughly $-4.46$ to $-4.18$.

**Match played (z):**

- Bundesliga: 0.1299  
- Ligue 1: 0.0945  
- Premier League: 0.1540  
- La Liga: 0.1029  
- Serie A: 0.1027  

All positive, all highly significant.

**Season (z):**

- Bundesliga: 0.5841  
- Ligue 1: 0.6314  
- Premier League: 0.7199  
- La Liga: 0.4936  
- Serie A: 0.5777  

Again, all positive and highly significant, but with noticeable variation across leagues.

### 3.2 Interpretation

- **Within each league**, more matches and later seasons are consistently associated with **higher injury rates**.
- The **magnitude** of the season effect differs:
  - Strongest in the **Premier League** (~0.72),
  - Weakest in **La Liga** (~0.49),
  - Others in between (~0.58–0.63).

So the **direction of effect is stable across leagues**, but the **strength** of the time trend and baseline level varies.

---

## 4. Meta-analysis of league-specific coefficients

You then treat each league’s coefficient as a “study” and perform meta-analyses (k = 5) for:

- Intercept  
- `match_played_z`  
- `season_z`  

using both **fixed-effect** and **random-effects** models.

### 4.1 Intercept

#### Fixed-effect model

- Estimate: $-4.3017$  
- SE: 0.0173  
- 95% CI: $[-4.3355, -4.2679]$  
- Heterogeneity:  
  - $I^2 = 91.43\%$ → very high  
  - Q-test p \< 0.0001

Interpretation: the **average baseline log injury rate** across leagues is about $-4.30$, but there is **substantial between-league heterogeneity** in intercepts.

#### Random-effects model

- Estimate: $-4.2930$  
- SE: 0.0579  
- 95% CI: $[-4.4064, -4.1795]$  
- Predictive interval: $[-4.56, -4.03]$  
- $I^2 = 91.01\%$, $\tau \approx 0.123$

Converting the pooled intercept:

$$
e^{-4.2930} \approx 0.0137
$$

So roughly **14 injuries per 1,000 exposure units** on average, with meaningful variation across leagues.

**Plain language:**  
The **average baseline injury rate** across the five leagues is similar, but the high $I^2$ tells us that **leagues differ more than we’d expect by chance**. The random-effects model acknowledges that and gives a wider interval and a predictive range for a “new” league.

---

### 4.2 `match_played_z`

#### Fixed-effect model

- Estimate: 0.1132  
- SE: 0.0089  
- 95% CI: $[0.0958, 0.1306]$  
- $I^2 = 27.37\%$, Q p = 0.2391 (no strong evidence of heterogeneity)

#### Random-effects model

- Estimate: 0.1140  
- SE: 0.0102  
- 95% CI: $[0.0940, 0.1339]$  
- Predictive interval: $[0.0846, 0.1433]$  
- $I^2 = 23.26\%$, very low $\tau^2$

Exponentiating:

$$
\text{IRR}_{\text{match, pooled}} = e^{0.1140} \approx 1.12
$$

So a **1 SD increase in matches** is associated with about a **12% higher injury rate**, and this effect is **remarkably consistent across leagues**.

**Plain language:**  
The relationship between **number of matches and injury risk** is **stable and robust** across all five leagues. More matches → more injuries, and the size of that effect is similar everywhere.

---

### 4.3 `season_z`

#### Fixed-effect model

- Estimate: 0.5901  
- SE: 0.0088  
- 95% CI: $[0.5728, 0.6074]$  
- $I^2 = 93.77\%$, Q p \< 0.0001 → very high heterogeneity

#### Random-effects model

- Estimate: 0.6006  
- SE: 0.0368  
- 95% CI: $[0.5284, 0.6727]$  
- Predictive interval: $[0.4283, 0.7729]$  
- $I^2 = 94.18\%$, $\tau \approx 0.080$

Exponentiating:

$$
\text{IRR}_{\text{season, pooled}} = e^{0.6006} \approx 1.82
$$

So a **1 SD increase in season** (moving forward in time) is associated with about an **82% higher injury rate on average**, but the **strength of this trend varies a lot between leagues**.

**Plain language:**  
Injury rates are **rising over time in all leagues**, but **not equally**. Some leagues (e.g., Premier League) show a stronger increase than others (e.g., La Liga). The random-effects model captures this by giving a wider interval and a predictive range for what you might see in another league.

---

## 5. BLUPs and pooled predictions

The BLUPs (Best Linear Unbiased Predictions) give **league-specific predicted coefficients** under the random-effects meta-analytic framework.

### 5.1 Intercept BLUPs

- Range roughly from $-4.45$ (Premier League) to $-4.20$ (La Liga), with narrow prediction intervals per league.
- Pooled estimate: $-4.2930$ (as above).

**Interpretation:**  
Leagues differ modestly in baseline injury rates, but all sit in a fairly tight band around the pooled value.

### 5.2 `match_played_z` BLUPs

- League-specific predictions: ~0.108–0.122  
- Pooled: 0.1140, 95% CI $[0.0940, 0.1339]$

**Interpretation:**  
The **effect of match load is extremely consistent**—every league’s BLUP is close to the pooled estimate.

### 5.3 `season_z` BLUPs

- League-specific predictions:  
  - La Liga: ~0.50  
  - Bundesliga/Serie A: ~0.58  
  - Ligue 1: ~0.63  
  - Premier League: ~0.71  
- Pooled: 0.6006, 95% CI $[0.5284, 0.6727]$

**Interpretation:**  
All leagues show **positive time trends**, but the **Premier League sits at the high end**, La Liga at the low end, with others in between.

---

## 6. Practical interpretation and implications

### 6.1 Core findings

- **Injury rates are increasing over time** (large, consistent positive `season_z` effect). This is possibly affected by better coverage of injuries in more recent years.
- **More matches → more injuries**, with a **stable ~10–15% increase per SD of match load** across all leagues.
- **Baseline injury levels differ modestly between leagues**, with La Liga and Ligue 1 slightly lower than Bundesliga, and Premier League/Serie A similar to Bundesliga.
- **Heterogeneity is high for intercepts and time trends**, but **low for the match-load effect**.

### 6.2 What this means in football terms

- **Workload management:**  
  The match-load effect is robust: **adding more matches reliably increases injury risk**, regardless of league. This supports policies around squad rotation, fixture congestion management, and rest periods.

- **Temporal trend:**  
  The strong positive time trend suggests that **modern football is becoming more injury-prone**—possibly due to higher intensity, tighter schedules, or changes in style of play. The fact that this trend is stronger in some leagues (e.g., Premier League) hints at **league-specific structural or tactical factors**.

- **Cross-league comparisons:**  
  While leagues differ somewhat in baseline injury rates, the **mechanisms linking load and time to injuries are broadly similar**. This makes cross-league lessons and interventions more transferable, especially around match congestion.

---

## 7. Limitations and considerations

- **Poisson assumption:**  
  If there is overdispersion (variance > mean), a negative binomial model might be more appropriate. The mixed model and random effects help, but don’t fully guarantee perfect fit.
- **Standardization:**  
  Effects are in units of **standard deviations**, which is great for comparability but less intuitive for practitioners. Translating SDs into actual match counts per league would help communication.
- **Unmeasured confounders:**  
  No direct adjustment for training load, travel, schedule density, or rule changes—these may partly drive the time trend.
- **Exposure scaling:**  
  Interpretation in “per 1,000 exposure units” depends on how `match_exposure` is defined (minutes, hours, etc.). Making that explicit would sharpen the practical meaning.

---

## 8. Short, plain-language summary

Across the top five European leagues, teams that **play more matches** and seasons in **more recent years** have **higher match injury rates**. The effect of **match load** is consistent across leagues, while the **increase over time** is strong everywhere but varies in size—strongest in the Premier League, weakest in La Liga. Baseline injury levels differ somewhat between leagues, but not dramatically. Overall, the meta-analysis paints a clear picture: **modern football is getting more injury-intensive, and fixture congestion is a key, stable risk factor across competitions.**

If you’d like, we can next translate these log-scale effects into **concrete example scenarios** (e.g., “what happens if a team plays 5 extra matches?”) for a report or presentation.
